# One-sample tests

## Objectives
:::objectives
**Questions**

- When do I perform a one-sample test?
- Which one-sample tests are there and what are the assumptions?
- How do I interpret and present the results of the tests?

**Objectives**

- Set out your hypothesis for single sample continuous data
- Be able to summarise and visualise the data in R
- Understand and assess the underlying assumptions of the tests
- Perform a one-sample t-test and Wilcoxon signed-rank test in R
- Know which test is appropriate when
- Be able to interpret and report the results
:::

## Purpose and aim
These tests are used when we have a single sample of continuous data. It is used to find out if the sample came from a parent distribution with a given mean (or median). This essentially boils down to finding out if the sample mean (or median) is "close enough" to our hypothesised parent population mean (or median).
So, in the figure below, we could use these tests to see what the probability is that the sample of ten points comes from the distribution plotted above it i.e. a population with a mean of 20 mm.

```{r cs1-one-sample-intro, echo=FALSE}
x<-seq(0,40,0.05)
randx<-rnorm(10,21,5)
plot(x,dnorm(x,20,5),type='l',ylim=c(-0.015,0.08),ylab="probability density",xlab="length (mm)")
abline(h=0)
abline(v=20,col="red",lty=2)
segments(20,0,20,0.09,col='red')
stripchart(randx,add=T,at=-0.01,pch=4)
text(5,0.04,expression(atop("Hypothetical","Distribution")),cex=1.5)
text(5,-0.01,"Sample",cex=1.5)
```

## Choosing a test
There are two tests that we are going to look at in this situation; the one-sample t-test, and the one-sample Wilcoxon signed rank test. Both tests work on the sort of data that we’re considering here, but they both have different assumptions.

If your data is normally distributed, then a one-sample t-test is appropriate. If your data aren’t normally distributed, but their distribution is symmetric, and the sample size is small then a one-sample Wilcoxon signed rank test is more appropriate.


## One-sample t-test
For each statistical test we consider there will be five tasks. These will come back again and again, so pay extra close attention.

:::highlight
1. Setting out of the hypothesis
2. Summarise and visualisation of the data
3. Assessment of assumptions
4. Implementation of the statistical test
5. Interpreting the output and presentation of results
:::

We won’t always carry these out in exactly the same order, but we will always consider each of the five tasks for every test.

### Section commands
New commands used in this section:

| Function| Description|
|:- |:- |
|`t.test()`| Performs a one-sample t-test, Student's t-test and Welch's t-test in later sections.|
|`qqnorm()`| Plots a Q-Q plot for comparison with a normal distribution.|
|`qqline()`| Adds a comparison line to the Q-Q plot.|
|`shapiro.test()`| Performs a Shapiro-Wilk test for normality.|

### Data and hypotheses
For example, suppose we measure the body lengths of male guppies (in mm) collected from the Guanapo River in Trinidad. We want to test whether the data support the hypothesis that the mean body is actually 20 mm. We form the following null and alternative hypotheses:

-	$H_0$: The mean body length is equal to 20mm ($\mu =$ 20).
-	$H_1$: The mean body length is not equal to 20mm ($\mu \neq$ 20).

We will use a one-sample, two-tailed t-test to see if we should reject the null hypothesis or not.

-	We use a **one-sample** test because we only have one sample.
-	We use a **two-tailed** t-test because we want to know if our data suggest that the true (population) mean is different from 20 mm in either direction rather than just to see if it is greater than or less than 20 mm (in which case we would use a one-tailed test).
-	We’re using a **t-test** because we don’t know any better yet and because I’m telling you to. We’ll look at what the precise assumptions/requirements need to be in a moment.

Make sure you have downloaded the data (see: [Datasets](#cs1-datasets)) and placed it in the `data/raw` folder within your working directory.

We then read in the data and create a vector containing the data.

```{r cs1-one-sample-read-data}
# import the data
fishlengthDF <- read.csv("data/raw/CS1-onesample.csv")

# create a vector containing the data
fishlength <- fishlengthDF$Guanapo
```

The first line reads the data into R and creates an object called a **data frame**. This data frame only contains a single column of numbers called "Guanapo" (the name of the river). In most situations, and for most statistical analyses, having our data stored in a data frame is exactly what we’d want. However, for one sample tests we actually need our data to be stored as a vector. So, the second line extracts the values that are in the Guanapo column of our `fishlengthDF` data frame and creates a simple vector of numbers that we have called `fishlength`. This step is only necessary for one-sample tests and when we look at more complex datasets, we won’t need to do this second step at all.

### Summarise and visualise
Summarise the data and visualise it:

```{r cs1-one-sample-summary}
summary(fishlength)

boxplot(fishlength, main = "Male guppies", ylab = "Length (mm)")
```

The data do not appear to contain any obvious errors, and whilst both the mean and median are less than 20 (`r round(mean(fishlength), digits = 1)` and `r round(median(fishlength), digits = 1)` respectively) it is not absolutely certain that the sample mean is sufficiently different from this value to be "statistically significant", although we may anticipate such a result.

### Implement the test
Perform a one-sample, two-tailed t-test:

```{r cs1-one-sample-t-test}
t.test(fishlength, mu = 20, alternative = "two.sided")
```

-	The first argument must be a numerical vector of data values.
-	The second argument must be a number and is the mean to be tested under the null hypothesis.
-	The third argument gives the type of alternative hypothesis and must be one of `two.sided`, `greater` or `less.` 

### Interpreting the output and report results
This is the output that you should now see in the console window:

```{r cs1-one-sample-t-test-result, echo=FALSE}
t.test(fishlength, mu = 20, alternative = "two.sided")
```

-	The 1st line gives the name of the test and the 2nd line reminds you what the dataset was called
-	The 3rd line contains the three key outputs from the test:
    *	The calculated t-value is -3.5492 (we’ll need this for reporting)
    *	There are 28 degrees of freedom (again we’ll need this for reporting)
    * The p-value is 0.001387. 
-	The 4th line simply states the alternative hypothesis
-	The 5th and 6th lines give the 95th confidence interval (we don’t need to know this)
-	The 7th, 8th and 9th lines give the sample mean again (18.29655).

The p-value on the 3rd line is what we’re most interested in. It gives the probability of us getting a sample such as ours if the null hypothesis were actually true.

So:

-	a high p-value means that there is a high probability of observing a sample such as ours and the null hypothesis is probably true whereas
-	a low p-value means that there is a low probability of observing a sample such as ours and the null hypothesis is probably not true.

It is important to realise that the p-value is just an indication and there is no absolute certainty here in this interpretation.

People, however like more definite answers and so we pick an artificial probability threshold (called a significance level) in order to be able to say something more decisive. The standard significance level is 0.05 and since our p-value is smaller than this we choose to say that "it is very unlikely that we would have this particular sample if the null hypothesis were true". Consequently, we can reject our null hypothesis and state that:

> A one-sample t-test indicated that the mean body length of male guppies ($\mu$ = 18.29mm) differs significantly from 20 mm (t = -3.55, df = 28, p = 0.0014).

The above sentence is an adequate concluding statement for this test and is what we would write in any paper or report. Note that we have included (in brackets) information on the actual mean value of our group($\mu$ = 18.29mm), the test statistic (t = -3.55), the degrees of freedom (df = 28), and the p-value (p = 0.0014). In some journals you are only required to report whether the p-value is less than the critical value (e.g. p < 0.05) but I would always recommend reporting the actual p-value obtained.

Please feel free to ask a demonstrator if any aspect of this section is unclear as this does form the core of classical hypothesis testing and the logic here applies to all of the rest of the tests.

### Assumptions
In order to use a t-test for this analysis (and for the results to be strictly valid) we have to make two assumptions:

1.	The parent distribution from which the sample is taken is normally distributed (and as such the sample data are normally distributed themselves).

:::note
It is worth noting though that the t-test is actually pretty robust in situations where the sample data are not normal. For sufficiently large sample sizes (your guess is as good as mine, but conventionally this means about 30 data points), you can use a t-test without worrying about whether the underlying population is normally distributed or not.
:::
2.	Each data point in the sample is independent of the others. This is in general not something that can be tested for and instead has to be considered from the sampling procedure. For example, taking repeated measurements from the same individual would generate data that are not independent.

The second point we know nothing about and so we ignore it here (this is an issue that needs to be considered from the experimental design), whereas the first assumption can be checked.
There are three ways of checking for normality:

In increasing order of rigour, we have 

1.	Histogram
2.	Quantile-quantile plot
3.	Shapiro-Wilk test

### Histogram of the data
Plot a histogram of the data, which gives:

```{r cs1-one-sample-t-test-histogram}
hist(fishlength, breaks = 15)
```

The distribution appears to be uni-modal and symmetric, and so it isn’t obviously non-normal. However, there are a lot of distributions that have these simple properties but which aren’t normal, so this isn’t exactly rigorous. Thankfully there are other, more rigorous tests.

NB. By even looking at this distribution to assess the assumption of normality we are already going far beyond what anyone else ever does. Nevertheless, we will continue.

### Q-Q plot of the data
Q-Q plot is the short for quantile-quantile plot. This **diagnostic** plot (as it is sometimes called) is a way of comparing two distributions. How Q-Q plots work won’t be explained here but ask a demonstrator if you really want to know what is going on.

Construct a Q-Q Plot of the quantiles of the data against the quantiles of a normal distribution:

```{r cs1-one-sample-t-test-qqplot}
# plot the Q-Q plot
qqnorm(fishlength)

# and add a comparison line
qqline(fishlength)
```

What is important to know is that if the data were normally distributed then all of the points should lie on (or close to) the diagonal line in this graph.

In this case, the points lie quite close to the line for the most part but the sample quantiles (points) from either end of the sample distribution are either smaller (below the line on the left) or larger (above the line on the right) than expected if they were supposed to be normally distributed. This suggests that the sample distribution is a bit more spread out than would be expected if it came from a normal distribution.

It is important to recognise that there isn’t a simple unambiguous answer when interpreting these types of graph, in terms of whether the assumption of normality has been well met or not and instead it often boils down to a matter of experience.

It is a very rare situation indeed where the assumptions necessary for a test will be met unequivocally and a certain degree of personal interpretation is always needed. Here you have to ask yourself whether the data are normal “enough” for you to be confident in the validity of the test.

Below are four examples of QQ plots for different types of distributions:

```{r cs1-one-sample-qqplot-normal-dist, echo=FALSE}
normalDat<-rnorm(200,0,1)
par(mfrow=c(1,2))
hist(normalDat,breaks=10 , freq=FALSE,main="Normal Distribution",xlab="")
qqnorm(normalDat)
qqline(normalDat)
```

These two graphs relate to 200 data points that have been drawn from a normal distribution. Even here you can see that the points do not all lie perfectly on the diagonal line in the QQ plot, and a certain amount of deviation at the top and bottom of the graph can happen just by chance (if I were to draw a different set of point then the graph would look slightly different).

```{r cs1-one-sample-qqplot-uniform-dist, echo=FALSE}
uniformDat<-runif(200,-1,1)
par(mfrow=c(1,2))
hist(uniformDat,breaks=10 , freq=FALSE,main="Uniform Distribution",xlab="")
qqnorm(uniformDat)
qqline(uniformDat)
```

These two graphs relate to 200 data points that have been drawn from a uniform distribution. Uniform distributions are more condensed than normal distributions, and this is reflected in the QQ plot having a very pronounced S-shaped pattern to it (this is colloquially known as **snaking**).

```{r cs1-one-sample-qqplot-t-dist, echo=FALSE}
tDat<-rt(200,1)
par(mfrow=c(1,2))
tDat<-tDat[which(abs(tDat)<20)]
hist(tDat,breaks=20 , freq = FALSE,main="T Distribution",xlab="")
qqnorm(tDat)
qqline(tDat)
```

These two graphs relate to 200 data points that have been drawn from a t distribution. t distributions are more spread out than normal distributions, and this is reflected in the QQ plot again having a very pronounced S-shaped pattern to it, but this time the snaking is a reflection of that observed for the uniform distribution.

```{r cs1-one-sample-qqplot-exponential-dist, echo=FALSE}
expDat<-rexp(200)
par(mfrow=c(1,2))
hist(expDat,breaks=10 , freq=FALSE,main="Exponential Distribution",xlab="")
qqnorm(expDat)
qqline(expDat)
```

These two graphs relate to 200 data points that have been drawn from an exponential distribution. Exponential distributions are not symmetric and are very skewed compared with normal distributions. The significant right-skew in this distribution is reflected in the QQ plot again having points that curve away above the diagonal line at both ends (a left-skew would have the points being below the line at both ends).

In all four cases it is worth noting that the deviations are only at the ends of the plot.

### Shapiro-Wilk test
This is one of a number of formal statistical test that assess whether a given sample of numbers come from a normal distribution. It calculates the probability of getting the sample data if the underlying distribution is in fact normal. It is very easy to carry out in R.

Perform a Shapiro-Wilk test on the data:

```{r cs1-one-sample-shapiro-test}
shapiro.test(fishlength)
```

-	The 1st line gives the name of the test and the 2nd line reminds you what the dataset was called
-	The 3rd line contains the two key outputs from the test:
    -	The calculated w-value is 0.9494 (we don’t need to know this)
    -	The p-value is 0.1764 

As the p-value is bigger than 0.05 (say) then we can say that there is insufficient evidence to reject the null hypothesis that the sample came from a normal distribution.

It is important to recognise that the Shapiro-Wilk test is not without limitations. It is rather sensitive to the sample size being considered. In general, for small sample sizes, the test is very relaxed about normality (and nearly all datasets are considered normal), whereas for large sample sizes the test can be overly strict, and it can fail to recognise datasets that are very nearly normal indeed.

### Assumptions overview
:::highlight
In terms of assessing the assumptions of a test it is always worth considering several methods, both graphical and analytic, and not just relying on a single method.
:::

In the `fishlength` example, the graphical Q-Q plot analysis was not especially conclusive as there was some suggestion of snaking in the plots, but the Shapiro-Wilk test gave a non-significant p-value (0.1764). Putting these two together, along with the original histogram and the recognition that there were only 30 data points in the dataset I personally would be happy that the assumptions of the t-test were met well enough to trust the result of the t-test, but you may not be...

In which case we would consider an alternative test that has less stringent assumptions (but is less powerful): the [one-sample Wilcoxon signed rank test](#wilcoxon-signed-rank).

### Exercise
:::exercise

The following data are the dissolving times (in seconds) of a drug in agitated gastric juice:

`42.7, 43.4, 44.6, 45.1, 45.6, 45.9, 46.8, 47.6`

Do these results provide any evidence to suggest that dissolving time for this drug is different from 45 seconds?

1. Write down the null and alternative hypotheses.
2. Summarise and visualise the data and perform an appropriate one-sample t-test.
    - What can you say about the dissolving time? (what sentence would you use to report this)
3. Check the assumptions for the test.
    - Was the test valid?

<details><summary>Answer</summary>

**Hypotheses**
$H_0$ : mean = 45s

$H_1$ : mean $\neq$ 45s

**Create data, summarise and visualise**
```{r}
dissolving<-c(42.7 , 43.4 , 44.6 , 45.1 , 45.6 , 45.9 , 46.8 , 47.6)
summary(dissolving)
```

```{r}
hist(dissolving)
boxplot(dissolving)
```

There are only 8 data points, so the default histogram looks a bit rubbish / uninformative. Thankfully the box-plot is a bit more useful here. We can see:

1. There don't appear to be any major errors in data entry and there aren't any huge outliers
2. The median value in the box-plot (the thick black line) is pretty close to 45 and so I wouldn't be surprised if the mean of the data isn't significantly different from 45. We can confirm that by looking at the mean and median values that we calculated using the summary command from earlier.
3. The data appear to be symmetric, and so whilst we can't tell if they're normal they're a least not massively skewed.

**Carry out t-test**
```{r}
t.test(dissolving , mu=45 , alternative = "two.sided")
```

A one-sample t-test indicated that the mean dissolving time of the drug is not significantly different from 45s (t=0.366 , df=7 , p=0.725),

**Explore Assumptions**
Normality:

```{r}
shapiro.test(dissolving)
```
```{r}
qqnorm(dissolving)
qqline(dissolving)
```

* The Shapiro test  has a p-value of 0.964 which (given that it is bigger than 0.05) suggests that the data are normal enough.
* The qq-plot isn't perfect, with some deviation of the points away from the line but since the points aren't accelerating away from the line and, since we only have 8 points, we can claim, with some slight reservations, that the assumption of normality appears to be adequately well met.

Overall, we are somewhat confident that the assumption of normality is well-enough met for the t-test to be an appropriate method for analysing the data. Note the ridiculous number of caveats here and the slightly political/slippery language I'm using. This is intentional and reflects the ambiguous nature of assumption checking. This is an important approach to doing statistics that you need to embrace.

In reality, if I found myself in this situation I would also try doing a non-parametric test on the data (Wilcoxon-signed Rank test) and see whether I get the same conclusion about whether the median dissolving time differs from 45s. Technically, you don't know about the Wilcoxon test yet as you haven't done that section of the handout. Anyway, if I get the same conclusion then my confidence in the result of the test goes up considerably; it doesn't matter how well an assumption has been met , I get the same result. If on the other hand I get a completely different conclusion from carrying out the non-parametric test then all bets are off; I now have very little confidence in my test result as I don't know which one to believe (in the case that the assumptions of the test are a bit unclear). In this example a Wilcoxon test also gives us a non-significant result and so all is good.

</details>
:::

## One-sample Wilcoxon signed-rank test {#wilcoxon-signed-rank}
This test also considers a single sample, however for this test (in contrast to the one sample t-test) we don’t have to assume that the parent distribution is normally distributed. We do still need the parent distribution (and consequently the sample) to be symmetric though. In this test we look to see if the _median_ of the parent distributions differs significantly from a given hypothesised value (in contrast with the t-test that looks at the _mean_).

### Section commands
New commands used in this section:

| Function| Description|
|:- |:- |
|`wilcox.test()`| Performs a one-sample wilcoxon signed-rank test, Wilcoxon test and Mann-Whitney test in later sections.|

Again, we use the `fishlength` dataset. The one-sample Wilcoxon signed-rank test allows to see if the _median_ body length is different from a specified value. Here we want to test whether the data support the hypothesis that the median body is actually 20 mm. The following null and alternative hypotheses are very similar to those used for the one sample t-test:

-	$H_0$: The median body length is equal to 20 mm ($\mu =$ 20).
-	$H_1$: The median body length is not equal to 20 mm ($\mu \neq$ 20).

We will use a one-sample, two-tailed Wilcoxon signed-rank test to see if we should reject the null hypothesis or not.

### Summarise and visualise
We did this before in the previous section, nothing really should have changed between now and then (if it has then you’re not off to a good start on this practical!)

### Implement the test
Perform a one-sample, two-tailed Wilcoxon signed-rank test:

```{r cs1-one-sample-wilcoxon, results='hide', warning=FALSE}
wilcox.test(fishlength, mu = 20, alternative = "two.sided")
```

The syntax is identical to the one-sample t-test we carried out earlier.

-	The first argument must be a numerical vector of data values.
-	The second argument must be a number and is the median to be tested under the null hypothesis.
-	The third argument gives the type of alternative hypothesis and must be one of `two.sided`, `greater` or `less.` 

### Interpreting the output and report results
This is the output that you should now see in the console window

```{r cs1-one-sample-wilcoxon-output, echo=FALSE}
wilcox.test(fishlength, mu = 20, alternative = "two.sided")
```

-	The first two lines give a warning (not an error) message regarding the implementation of this test. This can be safely ignored in this case as the p-value is so small, but essentially, it’s letting you know that some of the data values are identical to each other. This is not supposed to happen as we should be dealing with continuous data for this test, but in practice it’s not something that we need to worry ourselves with.
- The 3rd line gives the name of the test and the 4th line reminds you what the dataset was called
-	The 5th line contains the two key outputs from the test:
    -	The calculated statistic is 67.5 (we’ll need this for reporting)
    -	The p-value is 0.001222. 
-	The 6th line simply states the alternative hypothesis

Again, the p-value is what we’re most interested in. It gives the probability of us getting a sample such as ours if the null hypothesis were actually true.
So, in this case since our p-value is less than 0.05 we can reject our null hypothesis and state that:

> A one-sample Wilcoxon signed-rank test indicated that the median body length of male guppies ($\mu$ = 18.8 mm) differs significantly from 20 mm (V = 67.5, n = 29, p = 0.0012).

The above sentence is an adequate concluding statement for this test and is what we would write in any paper or report. Note that we have included (in brackets) information on the median value of the group ($\mu$ = 18.8 mm), the test statistic (V = 67.5), the number of observations (n = 29), and the p-value (p = 0.0012).

### Assumptions
In order to use a one-sample Wilcoxon rank-sum test for this analysis (and for the results to be strictly valid) we have to make two assumptions:

1.	The parent distribution from which the sample is symmetric
2.	Each data point in the sample is independent of the others. This is the same as for the t-test and is a common feature of nearly all statistical tests. Lack of independence in your data is really tough to deal with (if not impossible) and a large part of proper experimental design is ensuring this.

Whilst there are formal statistical tests for symmetry we will opt for a simple visual inspection using both a boxplot and a histogram.

Plot a histogram and a boxplot of the data:

```{r cs1-one-sample-wilcoxon-assumptions1, eval=FALSE}
hist(fishlength, breaks = 10)

boxplot(fishlength)
```

You get the following plots:

```{r cs1-one-sample-wilcoxon-assumptions2, echo=FALSE}
par(mfrow=c(1,2))
hist(fishlength, breaks = 10)
boxplot(fishlength)
```

Here we can see that whilst the distribution isn’t perfectly symmetric, neither is it heavily skewed to the left or right and we can make the call that the distribution is _symmetric enough_ for us to be happy with the results of the test.

### Exercise
:::exercise
Performing a Wilcoxon signed-rank test:

1. Analyse the drug dataset from before using a one-sample Wilcoxon signed-rank test
2. Discuss with a (virtual) neighbour which of the two tests you feel is best suited to the data.
3. Does it matter in this case?

<details><summary>Answer</summary>

**Hypotheses**
$H_0$ : median = 45s

$H_1$ : median $\neq$ 45s

**Wilcoxon signed-rank test**
```{r}
wilcox.test(dissolving , mu=45 , alternative = "two.sided")
```

A one-sample Wilcoxon-signed rank test indicated that the median dissolving time of the drug is not significantly different from 45 s (V=22, n=8 , p=0.64)

**Assumptions**
From the box-plot from the previous exercise we already know that the data are symmetric enough for the test to be valid.

**Discussion**
In terms of choosing between the two test we can see that both meet their respective assumptions and so both tests are valid. In this case both tests also agree in terms of their conclusions i.e. that the average dissolving time (either mean or median) doesn't differ significantly from the proposed value of 45 s.

* So one answer would be that it doesn't matter which test you use.
* Another answer would be that you should pick the test that measures the quantity you're interested in i.e. if you care about medians then use the Wilcoxon test, whereas if you care about means then use the t-test.
* A final answer would be that, since both test are valid we would prefer to use the test with greater **power**. t-tests always have more power than Wilcoxon tests (as long as they're valid) and so we could report that one. (We'll talk about this in the last session but power is effectively the capacity of a test to detect a significant difference - so more power is better).

</details>

:::

## Key points
:::keypoints
- One-sample tests are used when you have a single sample of continuous data
- We can summarise the data using the `summary()` function and visualise with `boxplot()`
- The t-test assumes that the data are normally distributed and independent of each other
- The Wilcoxon signed-rank test does not assume a normal distribution, but does require independent samples
- The `t.test()` compares if the _mean_ of the parent distribution differs from a hypothesised value, whereas `wilcox.test()` compares the _median_.
- A good way of assessing assumptions is to visually check by looking at the distribution with `hist()` and quantile-quantile plots with `qqnorm()` and `qqline()`
:::
